---
title: "3D Plotting in R: Examples"
output: html_notebook
---

Plotting in 3D has many useful applications. This notebook will illustrate some techniques used to create 3D plots in `R`. There are numerous resources that can be used for this purpose. 

## Preliminaries

Load the packages used in this notebook. `packr` contains a sample data set, `tidyverse` is a collection of packages for data manipulation and visualization, `plot3D` is a package for creating 3D plots, `plot3Drgl` is a package to convert `plot3D` static plots into interactive plots, and `plotly` is a package for creating interactive plots:
```{r message=FALSE}
library(packr)
library(tidyverse)
library(plot3D)
library(plot3Drgl)
library(plotly)
```

Summon the data set:
```{r}
data(energy_and_emissions)
```

Quickly review the contents of the data set:
```{r}
summary(energy_and_emissions)
```

Some variables are absolute values (e.g., population) and others are rates (e.g., GDP per capita). Mutate the dataframe to obain variables for energy consumption and emissions that are rates:
```{r}
energy_and_emissions <- energy_and_emissions %>%
  mutate(GDP = GDPPC * Population,
         GDPPC = GDPPC / 100000,
         EPC = bblpd/Population, 
         CO2PC_1995 = CO2_1995/Population, 
         CO2PC_2005 = CO2_2005/Population, 
         CO2PC_2015 = CO2_2015/Population)
```

## Pseudo-3D plots

The first thing to note is that any plots rendered on a flat surface such as a page in a book or a computer screen are pseudo-3D in the sense that they only exist in 2D. That said, there are different ways of representing 3 and even higher dimensions in a 2D surface.

We will begin with some plots that are more pseudo-3D than others.

Perhaps the simplest way to represent a 3D plot in two dimensions is by _projecting_ one of the dimensions on the other two, effectively flattening it but in such a way that an aspect of the 3D is still legible from the plot.

Consider the `energy_and_emissions` data set. It contains several variables, including population GDPPC, oil consumption in barrels of oil per day, and $CO_2$ emissions (all converted to per capita measurements). A common visualization tool is the scatterplot. This figure, for example, is the scatterplot of GDPPC

```{r}
ggplot(data = energy_and_emissions, aes(x = GDPPC, y = CO2PC_1995)) + geom_point()
```


```{r}
ggplot(data = energy_and_emissions, aes(x = EPC, y = CO2PC_1995)) + geom_point()
```


```{r}
ggplot(data = energy_and_emissions, aes(x = GDPPC, y = EPC, color = CO2PC_1995)) + 
  geom_point() + 
  scale_color_viridis_c(direction = -1)
```

In this plot, we "flatten" the variable for emissions, projecting it on the 2D plane of GDP per capita and energy consumptions per capita, and represent the "height" by means of colors. We can further emphasize the "height" of the emissions variable (its implied projection on the z axis) by further manipulating the attributes of the geometric objects, for instance their size:
```{r}
ggplot(data = energy_and_emissions, aes(x = GDPPC, y = EPC, color = CO2PC_1995, size = CO2PC_1995, shape = Continent)) + 
  geom_point() + 
  scale_color_viridis_c(direction = -1)
```

With some imagination, you can maybe visualize in your mind the scatter of points on the flattened third-dimension (the z-axis).

## 3D plots with `plot3D`

The package `

For convenience, we will rename the variables:
```{r}
GDPPC <- energy_and_emissions$GDPPC
EPC <- energy_and_emissions$EPC
CO2PC_1995 <- energy_and_emissions$CO2PC_1995
```


```{r}
scatter3D(x = GDPPC, y = EPC, z = CO2PC_1995)
```

### Color palettes

There are preset palettes (the default is `jet.col(n)` with `n`, the number of colors to generate, set to 100). Alternatives include: `jet2.col(n)`. This palette is similar to `jet.col` but brighter, since it lacks the deeper blues and reds:
```{r}
scatter3D(x = GDPPC, y = EPC, z = CO2PC_1995,
          col = jet2.col(n = 100))
```

Palette `gg.col(n)` uses colors similar to those used in `ggplot2`:
```{r}
scatter3D(x = GDPPC, y = EPC, z = CO2PC_1995,
          col = gg.col(n = 50))
```

The function `ramp.col()` creates a sequence of colors by interpolation, and is based on two or three colors, for example:
```{r}
scatter3D(x = GDPPC, y = EPC, z = CO2PC_1995,
          col = ramp.col(col = c("green", "red"), n = 50, alpha = 1))
```

And with three colors:
```{r}
scatter3D(x = GDPPC, y = EPC, z = CO2PC_1995,
          col = ramp.col(col = c("blue", "yellow", "red"), n = 50, alpha = 1))
```

### Aspect of plot and annotations

The shape of the markers can be changed by means of the argument `pch`. Here is a [list of symbols](http://www.sthda.com/english/wiki/r-plot-pch-symbols-the-different-point-shapes-available-in-r) available for plotting, The size of the symbol is controled by the argument `cex`:
```{r}
scatter3D(x = GDPPC, y = EPC, z = CO2PC_1995, 
          pch = 0,
          cex = 1.5)
```

We can also add annotations, such as labels and a legend:
```{r}
scatter3D(x = GDPPC, y = EPC, z = CO2PC_1995, 
          pch = 20,
          cex = 1.5,
          xlab = "GDPPC", ylab = "EPC", zlab = "CO2",
          clab = c("CO2 Emissions", "per Capita", "(kilotonnes)"))
```

We can also anotate the points in the plot. Unlike `ggplot2` which is based on a grammar of graphics that adds layers in a convenient way, `plot3D` uses a different syntax, whereby new elements are created and then _added_ to the existing plot. In this example, annotations are added by means of function `text3D()` (notice the argument `add`):  
```{r}
scatter3D(x = GDPPC, y = EPC, z = CO2PC_1995, 
          pch = 20,
          cex = 1.5,
          labels = row.names(energy_and_emissions),
          xlab = "GDPPC", ylab = "EPC", zlab = "CO2",
          clab = c("CO2 Emissions", "per Capita", "(kilotonnes)"))

text3D(x = GDPPC, y = EPC, z = CO2PC_1995, 
       cex = 0.75,
       labels = energy_and_emissions$Continent,
       add = TRUE)
```

### Changing the perspective

Two arguments control the viewing direction: `theta` and `phi`. These two are angles, `theta` is the angle with respect to the azimuth, whereas `phi` is the colatitude. The default values are $\theta = 40$ and $\phi = 40$ (in degrees).

We can see the effect of changing these values. For instance, if we change the value of `theta`, the perspective rotates with respect to the z-axis:
```{r}
scatter3D(x = GDPPC, y = EPC, z = CO2PC_1995, 
          pch = 20,
          cex = 1.5,
          theta = 90,
          labels = row.names(energy_and_emissions),
          xlab = "GDPPC", ylab = "EPC", zlab = "CO2",
          clab = c("CO2 Emissions", "per Capita", "(kilotonnes)"))
```

If we change the value of `phi`, the perspective rotates with respect to the x-y plane:
```{r}
scatter3D(x = GDPPC, y = EPC, z = CO2PC_1995, 
          pch = 20,
          cex = 1.5,
          theta = 40,
          phi = 90,
          labels = row.names(energy_and_emissions),
          xlab = "GDPPC", ylab = "EPC", zlab = "CO2",
          clab = c("CO2 Emissions", "per Capita", "(kilotonnes)"))
```

### Interactive 3D plots with `plot3Drgl`

Changing the perspective of the plot is a useful way to develop deeper insights into the data, but the process of regenerating the plot after changing the angles for the perspective can be a tedious process, and slow, depending on the size of the dataframe. To overcome this, package `plot3Drgl` can be used to create interactive plots based on objects created with `plot3D`. The function `plotrgl()` will create an interactive version of the most recent `plot3D` object:
```{r eval=FALSE}
scatter_labels <- scatter3D(x = GDPPC, y = EPC, z = CO2PC_1995, 
          pch = 20,
          cex = 1.5,
          labels = row.names(energy_and_emissions),
          xlab = "GDPPC", ylab = "EPC", zlab = "CO2",
          clab = c("CO2 Emissions", "per Capita", "(kilotonnes)"))

text3D(x = GDPPC, y = EPC, z = CO2PC_1995, 
       cex = 0.75,
       labels = energy_and_emissions$Continent,
       add = TRUE)

plotrgl()
```

### Other 3D objects

There are other types of objects (besides points and text) that can be plotted in 3D using `plot3D`, including:

- 3D lines
- 3D ribbons
- 3D slices
- 3D surfaces

Compute the linear regression (z = ax + by + d)
```{r}
mod <- lm(CO2PC_1995 ~ GDPPC + EPC)
summary(mod)
```

Predict values on regular xy grid
```{r}
grid.lines = 40
GDPPC.pred <- seq(min(GDPPC), max(GDPPC), length.out = grid.lines)
EPC.pred <- seq(min(EPC), max(EPC), length.out = grid.lines)
xy <- expand.grid(GDPPC = GDPPC.pred, EPC = EPC.pred)
CO2.pred <- matrix(predict(mod, newdata = xy), 
                 nrow = grid.lines, ncol = grid.lines)
```

Fitted points for droplines to surface
```{r}
fitpoints <- predict(mod)
```

Scatter plot with regression plane
```{r}
scatter3D(x = GDPPC, y = EPC, z = CO2PC_1995, 
          pch = 20,
          cex = 1.5,
          theta = 30,
          phi = 20,
          labels = row.names(energy_and_emissions),
          xlab = "GDPPC", ylab = "EPC", zlab = "CO2",
          clab = c("CO2 Emissions", "per Capita", "(kilotonnes)"),
          surf = list(x = GDPPC.pred, 
                y = EPC.pred, 
                z = CO2.pred,  
                facets = NA, 
                fit = fitpoints))
```

```{r}
scatter3D(x = GDPPC, y = EPC, z = CO2PC_1995, 
          pch = 20,
          cex = 1.5,
          theta = 30,
          phi = 20,
          labels = row.names(energy_and_emissions),
          xlab = "GDPPC", ylab = "EPC", zlab = "CO2",
          clab = c("CO2 Emissions", "per Capita", "(kilotonnes)"),
          surf = list(x = GDPPC.pred, 
                y = EPC.pred, 
                z = CO2.pred,  
                facets = NA, 
                fit = fitpoints))

text3D(x = GDPPC, y = EPC, z = CO2PC_1995, 
       cex = 0.75,
       labels = energy_and_emissions$Country,
       add = TRUE)

plotrgl()
```

## Interactive 3D plots with `plotly`

The package `plotly` provides an `R` interface to the popular plotly.js library of plotting functions. This library has a number of attractive attributes:

1. It is inspired by the grammar of graphics, so it feels similar to `ggplot2`. Essentially, it provides layering structures for creating plots.

2. It supports pipe operations (i.e. `%>%`) which makes programming plots intuitive and easier to read than, say, `plot3D`, because the output of a function can be passed on (i.e., _piped_) to the next function.

3. It creates web plots that can be embedded in R Markdown and html, so it is possible to work with the plots within R Studio, and the interactivity is retained when the output document is html.

The following example is somewhat advanced, so I will explain the different parts.

First, we need a dataframe, in our case `energy_and_emissions`. Notice how the dataframe is _piped_ to the `plot_ly()` function.
```
energy_and_emissions %>%
  plot_ly()
```

The arguments of the `plot_ly()` function map variables in the dataframe to aspects in the plot. For instance, for a 3D plot we need to tell `plotly` what to plot on the x-axis, y-axis, and z-axis. The format for this is as a formula, using `~`:
```
energy_and_emissions %>%
  plot_ly(x = ~GDPPC, 
          y = ~EPC, 
          z = ~CO2PC_1995)
```

We can also specify other aspect of the plot, such as the _color_ of the markers, their _size_, for example:
```
energy_and_emissions %>%
  plot_ly(x = ~GDPPC, 
          y = ~EPC, 
          z = ~CO2PC_1995, 
          color = ~Continent, 
          colors = colors,
          size = ~CO2PC_1995, 
          sizes = c(5, 40))
```

In addition, we need to specify the _shape_ of the markers:
```
energy_and_emissions %>%
  plot_ly(x = ~GDPPC, 
          y = ~EPC, 
          z = ~CO2PC_1995,
          color = ~Continent, 
          colors = colors,
          size = ~CO2PC_1995, 
          sizes = c(5, 40),
          marker = list(symbol = 'circle', sizemode = 'diameter',
                        line = list(width = 2, color = '#000000')))
```

Finally, we can create a `text` attribute as follows:
```
energy_and_emissions %>%
  plot_ly(x = ~GDPPC, 
          y = ~EPC, 
          z = ~CO2PC_1995, 
          color = ~Continent, 
          colors = colors,
          size = ~CO2PC_1995, 
          sizes = c(5, 40),
          marker = list(symbol = 'circle', sizemode = 'diameter',
                        line = list(width = 2, color = '#000000')),
          text = ~paste('Country:', Country, 
                        '<br>GDP per capita:', round(GDPPC, 3), 
                        '<br>Energy Consumption:', round(EPC,3),
                        '<br>CO2 emmissions 1995:',  round(CO2PC_1995, 3)))
```

Putting this together:
```{r}
colors <- c('#4AC6B7', '#1972A4', '#965F8A', '#FF7070', '#C61951')

energy_and_emissions %>%
  plot_ly(x = ~GDPPC, 
          y = ~EPC, 
          z = ~CO2PC_1995, 
          color = ~Continent, 
          colors = colors,
          size = ~CO2PC_1995, 
          sizes = c(5, 40),
          marker = list(symbol = 'circle', sizemode = 'diameter',
                        line = list(width = 2, color = '#000000')),
          text = ~paste('Country:', Country, 
                        '<br>GDP per capita:', round(GDPPC, 3), 
                        '<br>Energy Consumption:', round(EPC,3),
                        '<br>CO2 emmissions 1995:',  round(CO2PC_1995, 3)))
```

The function creates a `plotly` object that can then be passed on to function `layout()` which is used to control other aspects of the plot, such as the title, the background of the plot, the background of the image, and the _scene_, basically a collection of attributes for different elements of the plot such as the x-axis, the y-axis, the z-axis, etc.
```{r message=FALSE, warning=FALSE}
colors <- c('#4AC6B7', '#1972A4', '#965F8A', '#FF7070', '#C61951')

energy_and_emissions %>%
  plot_ly(x = ~GDPPC, 
          y = ~EPC, 
          z = ~CO2PC_1995, 
          color = ~Continent, 
          colors = colors,
          size = ~CO2PC_1995, 
          sizes = c(5, 40),
          marker = list(symbol = 'circle', sizemode = 'diameter',
                        line = list(width = 2, color = '#000000')),
          text = ~paste('Country:', Country, 
                        '<br>GDP per capita:', round(GDPPC, 3), 
                        '<br>Energy Consumption:', round(EPC,3),
                        '<br>CO2 emmissions 1995:',  round(CO2PC_1995, 3))) %>%
  layout(title = 'GDP per capita v. Energy consumption per capita',
         scene = list(xaxis = list(title = 'GDP per capita',
                      gridcolor = 'rgb(255, 255, 255)',
                      zerolinewidth = 1,
                      ticklen = 5,
                      gridwidth = 2),
               yaxis = list(title = 'Energy Consumption',
                      gridcolor = 'rgb(255, 255, 255)',
                      zerolinewidth = 1,
                      ticklen = 5,
                      gridwith = 2),
               zaxis = list(title = 'CO2 Emissions',
                            gridcolor = 'rgb(255, 255, 255)',
                            zerolinewidth = 1,
                            ticklen = 5,
                            gridwith = 2)),
         paper_bgcolor = 'rgb(243, 243, 243)',
         plot_bgcolor = 'rgb(243, 243, 243)')
```

In `plotly` it is possible to add one more aspect to the plot by using the attribute `frame`. A frame is similar to a `facet` in `ggplot2`, that is, a slice of the data according to a categorical variable. An intuitive way to think of frames is in terms of time series, and then the sequence of plots generated become an animation of the time series. However, the frames could be any other attribute, although for the animations to be effective it helps if the variable for the frame is ordinal, so that there is clear "forward" and "backward" sense.
```{r}
co2_95to15 <- energy_and_emissions %>% 
  dplyr::select(Country, Continent, GDPPC, EPC, CO2PC_1995, CO2PC_2005, CO2PC_2015) %>% # First select relevant variables
  gather(Year, CO2, -c(Country, Continent, GDPPC, EPC)) %>% # Gather CO2 columns: it is important to exclude from this operation the columns Country and GDP
  mutate(Year = factor(Year, 
                       levels = c("CO2PC_1995", "CO2PC_2005", "CO2PC_2015"),
                       labels = c("1995", "2005", "2015"))) # Relabel the years
```

Notice the additional attribute `frame` below:
```{r message=FALSE, warning=FALSE}
plot_ly(co2_95to15, x = ~GDPPC, y = ~EPC, z = ~CO2, 
        color = ~Continent, 
        size = ~CO2, 
        colors = colors,
        frame = ~Year,
        marker = list(symbol = 'circle', sizemode = 'diameter',
                      line = list(width = 2, color = '#000000')), 
                      sizes = c(5, 40),
        text = ~paste('Country:', Country, 
                      '<br>GDP per capita:', round(GDPPC, 3), 
                      '<br>Energy Consumption:', round(EPC,3),
                      '<br>CO2 emmissions 1995:',  round(CO2PC_1995, 3))) %>%
  layout(title = 'GDP per capita v. Energy consumption per capita',
         scene = list(xaxis = list(title = 'GDP per capita',
                      gridcolor = 'rgb(255, 255, 255)',
                      zerolinewidth = 1,
                      ticklen = 5,
                      gridwidth = 2),
               yaxis = list(title = 'Energy Consumption',
                      gridcolor = 'rgb(255, 255, 255)',
                      zerolinewidth = 1,
                      ticklen = 5,
                      gridwith = 2),
               zaxis = list(title = 'CO2 Emissions',
                            gridcolor = 'rgb(255, 255, 255)',
                            range = c(0, 0.1),
                            zerolinewidth = 1,
                            ticklen = 5,
                            gridwith = 2)),
         paper_bgcolor = 'rgb(243, 243, 243)',
         plot_bgcolor = 'rgb(243, 243, 243)') %>% 
  animation_opts(
    1300, redraw = FALSE
  )
```

### Total GDP and energy

The previous visualizations were all for per capita measurements. For the sake of comparison, we will repeat the visualizations but now for total GDP and total emissions.

Gather the columns with the $CO_2$ emissions data as for faceting:
```{r}
co2_95to15 <- energy_and_emissions %>% 
  dplyr::select(Country, Continent, GDP, bblpd, CO2_1995, CO2_2005, CO2_2015) %>% # First select relevant variables
  gather(Year, CO2, -c(Country, Continent, GDP, bblpd)) %>% # Gather CO2 columns: it is important to exclude from this operation the columns Country and GDP
  mutate(Year = factor(Year, 
                       levels = c("CO2_1995", "CO2_2005", "CO2_2015"),
                       labels = c("1995", "2005", "2015"))) # Relabel the years
```

Once the data has been tidyied, it is possible to add one dimension as a `frame` which enables animations:
```{r message=FALSE, warning=FALSE}
plot_ly(co2_95to15, x = ~GDP, y = ~bblpd, z = ~CO2, 
        color = ~Continent, 
        size = ~CO2, 
        colors = colors,
        frame = ~Year,
        marker = list(symbol = 'circle', sizemode = 'diameter',
                      line = list(width = 2, color = '#000000')), 
                      sizes = c(5, 40),
        text = ~paste('Country:', Country, 
                      '<br>GDP:', round(GDPPC, 3), 
                      '<br>Energy Consumption:', round(EPC,3),
                      '<br>CO2:',  round(CO2, 3))) %>%
  layout(title = 'GDP v. Energy consumption',
         scene = list(xaxis = list(title = 'GDP',
                      gridcolor = 'rgb(255, 255, 255)',
                      zerolinewidth = 1,
                      ticklen = 5,
                      gridwidth = 2),
               yaxis = list(title = 'Energy Consumption',
                      gridcolor = 'rgb(255, 255, 255)',
                      zerolinewidth = 1,
                      ticklen = 5,
                      gridwith = 2),
               zaxis = list(title = 'CO2 Emissions',
                            gridcolor = 'rgb(255, 255, 255)',
                            range = c(0, 11e6),
                            zerolinewidth = 1,
                            ticklen = 5,
                            gridwith = 2)),
         paper_bgcolor = 'rgb(243, 243, 243)',
         plot_bgcolor = 'rgb(243, 243, 243)') %>% 
  animation_opts(
    1300, redraw = FALSE
  )
```

## Concluding remarks

This note presented some tools for displaying multi-dimensional information in pseudo-3D plots. All packages "flatten" the information in some way for representation in two-dimensional surfaces, but using perspective and interactivity it is possible to get a more real 3D feeling. The addition of interactivity in `plot3Drgl` and `plotly` makes the exploration of multi-dimensional data more convenient, appealing, and effective. An advantage of plots created with `plotly` is that they are web objects that can be embedded in R Markdown as well as html, so they can be used effectively as presentation tools. Plots created with `plot3Drgl` in contrast are more difficult to embed in an output document. Furthermore, animations in `plotly` increase the dimensionality of the representation by one, producing pseudo-4D plots.

There are many other posibilities with 3D visualization, and I would encourage you to check the suggested readings for ideas.

## Suggested readings

[`ggplot2`: A package for a grammar of graphics]()  
[`plot3D`: A package for plotting multidimensional data](https://CRAN.R-project.org/package=plot3D)  
[`plotly`: A package for creating interactive web graphics via plotly.js](https://CRAN.R-project.org/package=plotly)