---
title: "packr-vignette"
output: rmarkdown::html_vignette
abstract: An important element of reproducible research is documenting all steps of the process, including data analysis. This vignette, developed for the `packr` package (a minimal example of package building), will show how to document data analysis in a self-contained and reproducible document. To this end, the data set `energy_and_emissions` (included in the `packr` package) will be used.
vignette: >
  %\VignetteIndexEntry{packr-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction


This vignette illustrate the use of the dataset and utility functions included in the package `packr`.

## Knitr and chunk options

Words go here.

## Preliminaries

Load the packages used in this vignette:
```{r setup}
library(packr)
library(tidyverse)
```

## Loading the data

To load the data, use the function `data()`:
```{r load-data}
data("energy_and_emmisions")
```

To inspect the dataframe, use the function `summary()`
```{r summary-data}
summary(energy_and_emissions)
```

The data frame consists of 10 variables. The variable definitions can be consulted in the help file:
```{r call-help}
?energy_and_emissions
```

## What is the per capita consumption of oil by country?

The dataframe includes information on population, GDP per capita, energy consumption, and emissions for world countries. The consumption of energy (in barrels per day) is for the country. But what is the consumption per capita? We need to divide the variable `bblpd` by `Population`. Use `dplyr::mutate()` to add this variable to the dataframe:
```{r energy-per-capita}
energy_and_emissions <- energy_and_emissions %>%
  mutate(EPC = bblpd/Population)
```

Check the descriptive statistics of `EPC` (energy consumption in barrels per day per person):
```{r}
summary(energy_and_emissions$EPC)
```

The maximum consumption is approximately `r round(max(energy_and_emissions$EPC), 2)` barrels per person per day. Which country is that?
```{r country-max-epc}
energy_and_emissions[energy_and_emissions$EPC == max(energy_and_emissions$EPC), "Country"]
```

The country with the highest per capita oil consumption in the world according the the data is Singapore.

## Are GDP per capita and energy consumption per capita related?

To answer this question, we can create a scatterplot of the two variables:
```{r}
ggplot(data = energy_and_emissions, aes(x = GDPPC, y = EPC)) +
  geom_point()
```

Calculate the correlation between these two variables:
```{r}
cor(energy_and_emissions$GDPPC, energy_and_emissions$EPC)
```

There is a moderately strong correlation between these two variables.
