---
title: An Example of a Term Paper
author:
  - name: Antonio Paez
    email: paezha@mcmaster.ca
    affiliation: School of Geography and Earth Sciences
    footnote: Corresponding Author
address:
  - code: School of Geography and Earth Sciences
    address: McMaster University, 1280 Main St W, Hamilton, ON, L8S 4K1 Canada
abstract: |
  This is an example of a paper created using the package `rticles`. It is an example of the final deliverable in the course GEO 712 Reproducible Research Workflow. The objective of the paper is to put into practice skills learned in the course, in the form of a self-contained, reproducible document in the format of a journal article.

  Notice that for this paper, the instructors will not evaluate the research per se, but rather the reproducibility of the research.

journal: "A Journal of Imaginary Results"
date: "`r Sys.Date()`"
bibliography: mybibfile.bib
#linenumbers: true
#numbersections: true
csl: elsevier-harvard.csl
output: rticles::elsevier_article
---

```{r load-packages, include = FALSE}

# Run only if you need to install `packr`
#devtools::install_github("paezha/Reproducible-Research-Workflow/Session-07-Creating-R-Packages-and-Documenting-Data/packr", build_vignettes = TRUE)

library(packr)
library(tidyverse)
library(gridExtra)
library(kableExtra)
```

```{r load-data, include = FALSE}
data("energy_and_emissions")
```

```{r data-preparation, include=FALSE}
# Use `dplyr::mutate` to create two new variables: GDP, obtained as the product of GDP per capita times the population, and ECP, the energy consumption per capita, obtained as the ratio of bblpd to population

energy_and_emissions <- energy_and_emissions %>%
  mutate(GDP = GDPPC * Population, EPC = bblpd / Population)
```

Introduction
============

The economy of a nation is tied to its consumption of energy, since every process of production requires energy as an input [@Warr2010energy]. However, the strength of the relationship between the economy and the consumption of energy varies. Some countries were more successful than others in terms of decoupling their productive processes from energy after the oil shocks of the 1970s [@Mackillop1989decoupling]. This was achieved by increasing the efficiency of production, so that the same output could be produced using less energy, or in somewhat different terms, by improving their energy intensity.

The relationship between economic output and energy consumption is of interest at a time when the effects of a carbon-intense economy is creating a heavy environmental burden. A relevant question is, what countries are more energy-efficient, and can we learn from them. To explore this question we will consider data on national energy use (in barrels of oil per day), economic output (GDP), and $CO_2$ emissions. 

Data Analysis
============

Data were collected from a variety of sources as documented in the package `packr` (see [paezha/Reproducible-Research-Workflow/Session-07-Creating-R-Packages-and-Documenting-Data/packr]). The summary statistics of the data appear in Table \ref{tab:descriptive-statistics}. Very large disparities can be observed in terms of all indicators of interest: while some countries have a GDP per capita measures in hundreds of dollars, others measure their GDP per capita in tens of thousands of dollars. Similarly, the difference in consumption of oil between the country with the lowest to the highest use of this resource is three orders of magnitude higher. The table also illustrates how emissions of $CO_2$ for the worst polluters increased by a factor of two between 1995 and 2015; emissions at the bottom, in contrast, increased by a factor of approximately 1.5 in the same period of time.

```{r descriptive-statistics-1, include=FALSE}
descriptive.df <- data.frame(Statistic = c("Mean", "Min", "Max", "Standard Deviation"),
                             Population = c(mean(energy_and_emissions$Population),
                                            min(energy_and_emissions$Population),
                                            max(energy_and_emissions$Population),
                                            sd(energy_and_emissions$Population)),
                             GDPPC = c(mean(energy_and_emissions$GDPPC),
                                       min(energy_and_emissions$GDPPC),
                                       max(energy_and_emissions$GDPPC),
                                       sd(energy_and_emissions$GDPPC)),
                             bblpd = c(mean(energy_and_emissions$bblpd),
                                       min(energy_and_emissions$bblpd),
                                       max(energy_and_emissions$bblpd),
                                       sd(energy_and_emissions$bblpd)),
                             CO2_1995 = c(mean(energy_and_emissions$CO2_1995),
                                       min(energy_and_emissions$CO2_1995),
                                       max(energy_and_emissions$CO2_1995),
                                       sd(energy_and_emissions$CO2_1995)),
                             CO2_2005 = c(mean(energy_and_emissions$CO2_2005),
                                       min(energy_and_emissions$CO2_2005),
                                       max(energy_and_emissions$CO2_2005),
                                       sd(energy_and_emissions$CO2_2005)),
                             CO2_2015 = c(mean(energy_and_emissions$CO2_2015),
                                       min(energy_and_emissions$CO2_2015),
                                       max(energy_and_emissions$CO2_2015),
                                       sd(energy_and_emissions$CO2_2015)))

kable(descriptive.df)
```

```{r descriptive-statistics-2, include=FALSE}
descriptive.df <- data.frame(Statistic = c("Mean", "Min", "Max", "Standard Deviation"),
                             Population = c(mean(energy_and_emissions$Population),
                                            min(energy_and_emissions$Population),
                                            max(energy_and_emissions$Population),
                                            sd(energy_and_emissions$Population)),
                             GDPPC = c(mean(energy_and_emissions$GDPPC),
                                       min(energy_and_emissions$GDPPC),
                                       max(energy_and_emissions$GDPPC),
                                       sd(energy_and_emissions$GDPPC)),
                             bblpd = c(mean(energy_and_emissions$bblpd),
                                       min(energy_and_emissions$bblpd),
                                       max(energy_and_emissions$bblpd),
                                       sd(energy_and_emissions$bblpd)),
                             CO2_1995 = c(mean(energy_and_emissions$CO2_1995),
                                       min(energy_and_emissions$CO2_1995),
                                       max(energy_and_emissions$CO2_1995),
                                       sd(energy_and_emissions$CO2_1995)),
                             CO2_2005 = c(mean(energy_and_emissions$CO2_2005),
                                       min(energy_and_emissions$CO2_2005),
                                       max(energy_and_emissions$CO2_2005),
                                       sd(energy_and_emissions$CO2_2005)),
                             CO2_2015 = c(mean(energy_and_emissions$CO2_2015),
                                       min(energy_and_emissions$CO2_2015),
                                       max(energy_and_emissions$CO2_2015),
                                       sd(energy_and_emissions$CO2_2015)))

kable(descriptive.df) %>%
  kable_styling(latex_options = c("scale_down"))
```

```{r descriptive-statistics-3, include=FALSE}
descriptive.df <- data.frame(Statistic = c("Mean", "Min", "Max", "Standard Deviation"),
                             Population = c(mean(energy_and_emissions$Population/1e6),
                                            min(energy_and_emissions$Population/1e6),
                                            max(energy_and_emissions$Population/1e6),
                                            sd(energy_and_emissions$Population/1e6)),
                             GDPPC = c(mean(energy_and_emissions$GDPPC),
                                       min(energy_and_emissions$GDPPC),
                                       max(energy_and_emissions$GDPPC),
                                       sd(energy_and_emissions$GDPPC)),
                             bblpd = c(mean(energy_and_emissions$bblpd/1e6),
                                       min(energy_and_emissions$bblpd/1e6),
                                       max(energy_and_emissions$bblpd/1e6),
                                       sd(energy_and_emissions$bblpd/1e6)),
                             CO2_1995 = c(mean(energy_and_emissions$CO2_1995/1e6),
                                       min(energy_and_emissions$CO2_1995/1e6),
                                       max(energy_and_emissions$CO2_1995/1e6),
                                       sd(energy_and_emissions$CO2_1995/1e6)),
                             CO2_2005 = c(mean(energy_and_emissions$CO2_2005/1e6),
                                       min(energy_and_emissions$CO2_2005/1e6),
                                       max(energy_and_emissions$CO2_2005/1e6),
                                       sd(energy_and_emissions$CO2_2005/1e6)),
                             CO2_2015 = c(mean(energy_and_emissions$CO2_2015/1e6),
                                       min(energy_and_emissions$CO2_2015/1e6),
                                       max(energy_and_emissions$CO2_2015/1e6),
                                       sd(energy_and_emissions$CO2_2015/1e6)))

kable(descriptive.df) %>%
  kable_styling(latex_options = c("scale_down"))
```

```{r descriptive-statistics-4, include=FALSE}
descriptive.df <- data.frame(Statistic = c("Mean", "Min", "Max", "Standard Deviation"),
                             Population = c(mean(energy_and_emissions$Population/1e6),
                                            min(energy_and_emissions$Population/1e6),
                                            max(energy_and_emissions$Population/1e6),
                                            sd(energy_and_emissions$Population/1e6)),
                             GDPPC = c(mean(energy_and_emissions$GDPPC),
                                       min(energy_and_emissions$GDPPC),
                                       max(energy_and_emissions$GDPPC),
                                       sd(energy_and_emissions$GDPPC)),
                             bblpd = c(mean(energy_and_emissions$bblpd/1e6),
                                       min(energy_and_emissions$bblpd/1e6),
                                       max(energy_and_emissions$bblpd/1e6),
                                       sd(energy_and_emissions$bblpd/1e6)),
                             CO2_1995 = c(mean(energy_and_emissions$CO2_1995/1e6),
                                       min(energy_and_emissions$CO2_1995/1e6),
                                       max(energy_and_emissions$CO2_1995/1e6),
                                       sd(energy_and_emissions$CO2_1995/1e6)),
                             CO2_2005 = c(mean(energy_and_emissions$CO2_2005/1e6),
                                       min(energy_and_emissions$CO2_2005/1e6),
                                       max(energy_and_emissions$CO2_2005/1e6),
                                       sd(energy_and_emissions$CO2_2005/1e6)),
                             CO2_2015 = c(mean(energy_and_emissions$CO2_2015/1e6),
                                       min(energy_and_emissions$CO2_2015/1e6),
                                       max(energy_and_emissions$CO2_2015/1e6),
                                       sd(energy_and_emissions$CO2_2015/1e6)))

kable(descriptive.df,
      digits = 3) %>%
  kable_styling(latex_options = c("scale_down"))
```

```{r descriptive-statistics-5, include=FALSE}
descriptive.df <- data.frame(Statistic = c("Mean", "Min", "Max", "Standard Deviation"),
                             Population = c(mean(energy_and_emissions$Population/1e6),
                                            min(energy_and_emissions$Population/1e6),
                                            max(energy_and_emissions$Population/1e6),
                                            sd(energy_and_emissions$Population/1e6)),
                             GDPPC = c(mean(energy_and_emissions$GDPPC),
                                       min(energy_and_emissions$GDPPC),
                                       max(energy_and_emissions$GDPPC),
                                       sd(energy_and_emissions$GDPPC)),
                             bblpd = c(mean(energy_and_emissions$bblpd/1e6),
                                       min(energy_and_emissions$bblpd/1e6),
                                       max(energy_and_emissions$bblpd/1e6),
                                       sd(energy_and_emissions$bblpd/1e6)),
                             CO2_1995 = c(mean(energy_and_emissions$CO2_1995/1e6),
                                       min(energy_and_emissions$CO2_1995/1e6),
                                       max(energy_and_emissions$CO2_1995/1e6),
                                       sd(energy_and_emissions$CO2_1995/1e6)),
                             CO2_2005 = c(mean(energy_and_emissions$CO2_2005/1e6),
                                       min(energy_and_emissions$CO2_2005/1e6),
                                       max(energy_and_emissions$CO2_2005/1e6),
                                       sd(energy_and_emissions$CO2_2005/1e6)),
                             CO2_2015 = c(mean(energy_and_emissions$CO2_2015/1e6),
                                       min(energy_and_emissions$CO2_2015/1e6),
                                       max(energy_and_emissions$CO2_2015/1e6),
                                       sd(energy_and_emissions$CO2_2015/1e6)))

kable(descriptive.df,
      digits = 3,
      booktabs = TRUE) %>%
  kable_styling(latex_options = c("striped", "scale_down"))
```

```{r descriptive-statistics-6, include=FALSE}
descriptive.df <- data.frame(Statistic = c("Mean", "Min", "Max", "Standard Deviation"),
                             Population = c(mean(energy_and_emissions$Population/1e6),
                                            min(energy_and_emissions$Population/1e6),
                                            max(energy_and_emissions$Population/1e6),
                                            sd(energy_and_emissions$Population/1e6)),
                             GDPPC = c(mean(energy_and_emissions$GDPPC),
                                       min(energy_and_emissions$GDPPC),
                                       max(energy_and_emissions$GDPPC),
                                       sd(energy_and_emissions$GDPPC)),
                             bblpd = c(mean(energy_and_emissions$bblpd/1e6),
                                       min(energy_and_emissions$bblpd/1e6),
                                       max(energy_and_emissions$bblpd/1e6),
                                       sd(energy_and_emissions$bblpd/1e6)),
                             CO2_1995 = c(mean(energy_and_emissions$CO2_1995/1e6),
                                       min(energy_and_emissions$CO2_1995/1e6),
                                       max(energy_and_emissions$CO2_1995/1e6),
                                       sd(energy_and_emissions$CO2_1995/1e6)),
                             CO2_2005 = c(mean(energy_and_emissions$CO2_2005/1e6),
                                       min(energy_and_emissions$CO2_2005/1e6),
                                       max(energy_and_emissions$CO2_2005/1e6),
                                       sd(energy_and_emissions$CO2_2005/1e6)),
                             CO2_2015 = c(mean(energy_and_emissions$CO2_2015/1e6),
                                       min(energy_and_emissions$CO2_2015/1e6),
                                       max(energy_and_emissions$CO2_2015/1e6),
                                       sd(energy_and_emissions$CO2_2015/1e6)))

kable(descriptive.df,
      digits = 3,
      booktabs = TRUE,
      col.names = c("Statistic", 
                    "Population (in millions)", 
                    "GDP per capita", 
                    "Energy use (in millions of barrels per day)", 
                    "CO2 emissions (in millions, 1995)", 
                    "CO2 emissions (in millions, 1995)", 
                    "CO2 emissions (in millions, 1995)")) %>%
  kable_styling(latex_options = c("striped", "scale_down"))
```

```{r descriptive-statistics-7, echo=FALSE}
descriptive.df <- data.frame(Statistic = c("Mean", "Min", "Max", "Standard Deviation"),
                             Population = c(mean(energy_and_emissions$Population/1e6),
                                            min(energy_and_emissions$Population/1e6),
                                            max(energy_and_emissions$Population/1e6),
                                            sd(energy_and_emissions$Population/1e6)),
                             GDPPC = c(mean(energy_and_emissions$GDPPC),
                                       min(energy_and_emissions$GDPPC),
                                       max(energy_and_emissions$GDPPC),
                                       sd(energy_and_emissions$GDPPC)),
                             bblpd = c(mean(energy_and_emissions$bblpd/1e6),
                                       min(energy_and_emissions$bblpd/1e6),
                                       max(energy_and_emissions$bblpd/1e6),
                                       sd(energy_and_emissions$bblpd/1e6)),
                             CO2_1995 = c(mean(energy_and_emissions$CO2_1995/1e6),
                                       min(energy_and_emissions$CO2_1995/1e6),
                                       max(energy_and_emissions$CO2_1995/1e6),
                                       sd(energy_and_emissions$CO2_1995/1e6)),
                             CO2_2005 = c(mean(energy_and_emissions$CO2_2005/1e6),
                                       min(energy_and_emissions$CO2_2005/1e6),
                                       max(energy_and_emissions$CO2_2005/1e6),
                                       sd(energy_and_emissions$CO2_2005/1e6)),
                             CO2_2015 = c(mean(energy_and_emissions$CO2_2015/1e6),
                                       min(energy_and_emissions$CO2_2015/1e6),
                                       max(energy_and_emissions$CO2_2015/1e6),
                                       sd(energy_and_emissions$CO2_2015/1e6)))

kable(descriptive.df,
      digits = 3,
      booktabs = TRUE,
      escape = FALSE,
      col.names = linebreak(c("Statistic", 
                              "Population\n(millions)",
                              "GDP per capita\n(USD)",
                              "Energy use\n(millions of \nbarrels per day)", 
                              "CO2 1995\n(millions\nkilotonnes)", 
                              "CO2 2005\n(millions\nkilotonnes)", 
                              "CO2 2015\n(millions\nkilotonnes)"),
                            align = "c"), #Lower case "c" for center
      caption = "\\label{tab:descriptive-statistics} Descriptive statistics: energy and emissions of world countries") %>% 
  kable_styling(latex_options = c("striped", "scale_down"))
```

Figure \ref{fig:energy-to-gdp} is a scatterplot of energy consumption to GPD. It can be seen that in general, greater economic output is associated with greater consumption of energy. However, there are some important differences. If we fit a regression line to this relationship, the line would indicate the _expected_ economic output for a given level of energy consumption. Points below the line would use more energy for a lower level of economic output than expected, whereas points above the line would represent greater economic output than expected, given their energy consumption.

```{r fig-energy-to-gdp, echo = FALSE, fig.cap="\\label{fig:energy-to-gdp} The relationship between energy consumption and economic output by world countries"}
ggplot(data = energy_and_emissions, aes(x = GDP, y = bblpd)) +
  geom_point() +
  ggtitle("Energy and Economic Output") +
  xlab("Energy (bblpd)") +
  ylab("GDP")
```

The regression line is estimated as follows:
$$
GDP_i = \beta_0 + \beta_1\text{bblpd}_i + \epsilon_i
$$

This is a linear model and can be estimated using ordinary least squares. The scatterplot of energy to GDP with this line is shown in Figure \ref{fig:energy-to-gdp-with-line}. Clearly, some countries are more efficient than others in that they can produce more with less energy. The more a point deviates from the regression line, the more efficient that economy is.

```{r fig-energy-to-gdp-with-line, echo = FALSE, fig.cap="\\label{fig:energy-to-gdp-with-line} A regression line gives the expected values of GDP given energy consumption"}
ggplot(data = energy_and_emissions, aes(x = GDP, y = bblpd)) +
  geom_point() +
  geom_smooth(method = "lm") +
  ggtitle("Energy and Economic Output: Expected vs Observed") +
  xlab("Energy (bblpd)") +
  ylab("GDP")
```

>>(Note: figures can also be multi-panel plots. One ways of doing this is by using the package `gridExtra`, which creates _tables_ of figures. As an example, see Figure \ref{fig:right-left-panel-plot}, which reproduces Figures \ref{fig:energy-to-gdp} and \ref{fig:energy-to-gdp-with-line} but using a side-by-side paneling format.

```{r fig-right-left-panel-plot, echo = FALSE, fig.cap="\\label{fig:right-left-panel-plot} Two plots in a single figure; left panel is Figure 1 and right panel is Figure 2"}

# Recreate Figure 1
fig1 <- ggplot(data = energy_and_emissions, aes(x = GDP, y = bblpd)) +
  geom_point() +
  ggtitle("Energy and Economic Output") +
  xlab("Energy (bblpd)") +
  ylab("GDP")

# Recreate Figure 2
fig2 <- ggplot(data = energy_and_emissions, aes(x = GDP, y = bblpd)) +
  geom_point() +
  geom_smooth(method = "lm") +
  ggtitle("Energy and Economic Output: Expected vs Observed") +
  xlab("Energy (bblpd)") +
  ylab("GDP")

grid.arrange(fig1, fig2, ncol = 2)
```

>>A second example that uses `gridExtra` is Figure \ref{fig:top-bottom-panel-plot}, which shows Figures \ref{fig:energy-to-gdp} and \ref{fig:energy-to-gdp-with-line} as a two-panel figure with one column and two rows).

```{r fig-top-bottom-panel-plot, echo = FALSE, fig.cap="\\label{fig:top-bottom-panel-plot} Two plots in a single figure; top panel is Figure 1 and bottom panel is Figure 2"}

# I do not need to recreate the figures, since they already exist after chunk `fig-right-left-panel-plot`
grid.arrange(fig1, fig2, nrow = 2)
```

Another question of interest is the way energy is related to the emission of Greenhouse Gases (GHG), particularly $CO_2$. Figure \ref{fig:gdp-emissions-by-year} suggests that for some countries, emissions increased over time more than proportionally relative to GDP.

```{r sample-plot, include=FALSE}

# This plot is not included in the output document, its purpose is to illustrate one possible way of generating a multi-panel figure by creating multiple plots.

co2_1995 <- ggplot(data = energy_and_emissions, aes(x = GDP, y = CO2_1995)) +
  geom_point() +
  ggtitle("GDP and Emissions (1995)") +
  xlab("GDP") +
  ylab("CO_2 Emissions")

co2_2005 <- ggplot(data = energy_and_emissions, aes(x = GDP, y = CO2_2005)) +
  geom_point()  +
  ggtitle("GDP and Emissions (1995)") +
  xlab("GDP") +
  ylab("CO_2 Emissions")

co2_2015 <- ggplot(data = energy_and_emissions, aes(x = GDP, y = CO2_2015)) +
  geom_point() +
    ggtitle("GDP and Emissions (1995)") +
  xlab("GDP") +
  ylab("CO_2 Emissions")

grid.arrange(co2_1995, co2_2005, co2_2015, nrow = 3)
```

```{r gather-co2, include=FALSE}

# There is a lot of repetition in the code above. An alternative is to use faceting in `ggplot2`. Faceting requires that the data be organized in a certain way; in this case, there are three columns with CO_2 data, for three different years. The format of the table is wide. For faceting, we need all CO_2 data in a single column, and the year to be a variable. The format of the table needs to be changed to long. In this chunk I will gather all CO_2 variables into a single column.

co2_95to15 <- energy_and_emissions %>% 
  dplyr::select(Country, GDP, CO2_1995, CO2_2005, CO2_2015) %>% # First select relevant variables
  gather(Year, CO2, -c(Country, GDP)) %>% # Gather CO2 columns: it is important to exclude from this operation the columns Country and GDP
  mutate(Year = factor(Year, 
                       levels = c("CO2_1995", "CO2_2005", "CO2_2015"),
                       labels = c("1995", "2005", "2015"))) # Relabel the years
```

```{r fig-gdp-emissions-by-year, echo = FALSE, fig.cap="\\label{fig:gdp-emissions-by-year} CO_2 emissions versus GDP by year"}

# This creates a multipanel figure using faceting. Compared to the approach above where multiple figures are created, this simplifies the coding of the plots and ensure a consistent aspect across all plots (including among other things same plot extents)

ggplot(data = co2_95to15, aes(x = GDP, y = CO2)) +
  geom_point() +
  ggtitle("GDP and Emissions by Year") +
  xlab("GDP") +
  ylab("CO_2 Emissions") +
  facet_wrap(~ Year)
```

To quantify the change in emissions by economic output, we can estimate regression models as follows:
$$
CO2_{iy} = \beta GDP_{iy} + \epsilon_{iy}
$$
for country $i$ and years $y=1995, 2005, 2015$.

The estimated models are found in Table \ref{tab:gdp-emissions-model-results}. It is noteworthy that the amount of $CO_2$ per million dollars of economic output increased from an average of 0.2899 kilotonnes in 1995, to 0.3749 kilotonnes in 2005, to 0.4481 in 2015. As well, notice how the coefficient of determination $R^2$ declines over that period of time, which indicates that there amount of unexplained variance has increased as well. Whereas in 1995, approximately 94% of the variance in $CO_2$ emissions by world countries was explained by GDP, by 2015 less than 70% was. Could this be a result of increased decoupling of the economy from energy? This is a question for future research.

```{r gdp-emissions-models, echo=FALSE}

# Here, I estimate regression models with emissions as the dependent variable and GDP as the independent variable. Comparing the slopes can be informative of the pollution potential of economic output. Before estimating the model, I will scale both variables so that GDP is in million USD, then I will force the intercept to pass through zero.

mod_1995 <- lm(formula = CO2 ~ 0 + GDP, data = co2_95to15 %>% filter(Year == "1995") %>% mutate(GDP = GDP/1e6))
mod_2005 <- lm(formula = CO2 ~ 0 + GDP, data = co2_95to15 %>% filter(Year == "2005") %>% mutate(GDP = GDP/1e6))
mod_2015 <- lm(formula = CO2 ~ 0 + GDP, data = co2_95to15 %>% filter(Year == "2015") %>% mutate(GDP = GDP/1e6))
```

```{r table-emissions-models-results, echo=FALSE}
models.df <- data.frame(Estimate.1 = mod_1995$coefficients, 
                        pval.1 = paste(ifelse(summary(mod_1995)$coefficients[,4] > 0.0001, 
                                              round(summary(mod_1995)$coefficients[,4], 4), 
                                              "< 0.0001")),
                        Estimate.2 = mod_2005$coefficients, 
                        pval.2 = paste(ifelse(summary(mod_2005)$coefficients[,4] > 0.0001, 
                                              round(summary(mod_2005)$coefficients[,4], 4), 
                                              "< 0.0001")),
                        Estimate.3 = mod_2015$coefficients, 
                        pval.3 = paste(ifelse(summary(mod_2015)$coefficients[,4] > 0.0001, 
                                              round(summary(mod_2015)$coefficients[,4], 4), 
                                              "< 0.0001")))

kable(models.df,
      digits = 4,
      booktabs = TRUE,
      escape = FALSE,
      col.names = c("$\\beta$", "p-val", "$\\beta$", "p-val", "$\\beta$", "p-val"),
      caption = "\\label{tab:gdp-emissions-model-results} Regression results: Emissions by GDP by year") %>%
  add_header_above(c("Variable" = 1, 
                     "1995" = 2, 
                     "2005" = 2, 
                     "2015" = 2),
                   escape = FALSE) %>%
  footnote(c(paste("$R^2$ (1995) = ", as.character(round(summary(mod_1995)$r.squared, 2))),
             paste("$R^2$ (2005) = ", as.character(round(summary(mod_2005)$r.squared, 2))),
             paste("$R^2$ (2015) = ", as.character(round(summary(mod_2015)$r.squared, 2)))),
           escape = FALSE)
```

Conclusions
==========

This document is an example of a term paper, and illustrates the following features:

- An R Markdown document created using the package `rticles` and the template for Elsevier articles
- Mathematical notation in LaTeX
- Graphical output
- References, including bibliographical references using BibTeX
- Tables

The PDF file was produced from source files in R Markdown. The analysis presented in this paper is self-contained and reproducible. The source documents can be downloaded from the link below:

https://github.com/paezha/Reproducible-Research-Workflow/tree/master/Session-11-Rticles-Tables-and-Citations/Elsevier-Template

References {#references .unnumbered}
==========
